\textbf{ Interpretable Gaussian Processes for Stellar Variability\\ I: Static Surfaces }

### 2020-12-28 tex/ms.tex
more text

\xxx{TODO: Discuss the Python code here.}

@@

>>>>

We implement our interpretable GP in the open-source, user-friendly
\Python package \starryprocess, which can be installed via \textsf{pip}\xspace
or from source on GitHub\footnote{\url{https://github.com/rodluger/starry_process}}.
The code is thoroughly unit-tested and well documented, with examples on how
to use the GP in custom inference problems. Users can choose, among other
options, whether or not
to marginalize over inclination (\S\ref{sec:inclination}) and
whether or not to model a normalized
process (\S\ref{sec:baseline}). Users can also choose the spherical harmonic
degree of the expansion, although it is recommended to use $l_\mathrm{max} = 15$
(see below).
Users may compute the mean vector and covariance matrix in either the spherical
harmonic basis or the flux basis, or they may sample from it or use it to
compute marginal likelihoods.
Arbitrary order limb darkening is implemented following
\citet{Agol2020}.

The code was designed to maximize the speed and numerical stability of the
computation. Although most of the computations of the expectation integrals
(see Appendix~\ref{sec:integrals}) involve many layers of nested sums over
all spherical harmonic coefficients, these may be expressed as high-dimensional
tensor products, which can be evaluated extremely efficiently on modern
hardware. Many of the expressions can also be either pre-computed or
computed recursively. To maximize the speed of the algorithm, the code is
implemented in hybrid \cpp/\Python using the just-in-time compilation
capability of the \theano package. Since all equations derived here
have closed form expressions, these can be differentiated in a straightforward
and numerically stable manner, enabling the computation of backpropagated
gradients within \theano. As such, \starryprocess is designed to work
out-of-the box with \theano-based inference tools such as \pymcthree
for NUTS/HMC or ADVI sampling.

Figure~\ref{fig:speed} shows the computational scaling of the \Python
implementation of the algorithm for the case where we condition the GP
on a specific value of the inclination (blue) and the case where we marginalize
over inclination (orange). Both curves show the time in seconds to compute the
likelihood (averaged over many trials) as a function of the number of points $K$
in a single light curve. For $K \lesssim 100$, the computation time is constant
at $10-20$ ms for both algorithms. This is the approximate time (on a typical
modern laptop) taken to compute the GP covariance matrix given a set of
hyperparameters $\pmb{\theta}_\bullet$. For larger values of $K$, the
cost approaches a scaling of $K^{2.5}$, which is dominated by the factorization
of the covariance matrix and the solve operation to compute the likelihood.
The likelihood marginalized over inclination is only slightly slower to compute,
thanks to the tricks discussed in Appendix~\ref{sec:inc-speedup}.

Many modern GP packages \citep[e.g.,][]{Ambikasaran2015,ForemanMackey2017}
have significantly better asymptotic scalings, but these are usually due to
specific structure imposed on the kernel functions, such as the assumption
of stationarity. Our kernel structure is determined by the physics (or perhaps
more accurately, the geometry) of stellar surfaces, and its nonstationarity
is a consequence of the normalization step in relative photometry. Moreover,
and unlike the kernels listed in \S\ref{sec:intro}, our
kernel is a nontrivial function of the hyperparameters
$\pmb{\theta}_\bullet$, so its computation is necessarily more expensive.
Nevertheless, the fact that our GP may be used for likelihood evaluation
in a small fraction of a second for typical datasets ($K \sim 1{,000}$)
makes it extremely useful for inference. Recall that we are implicitly
marginalizing over all of the properties of \emph{every spot} on the surface of
the star.

In ensemble analyses, we must compute the likelihood of each of the $M$ light curves
conditioned on $\pmb{\theta}_\bullet$. In practice, each star will have a
different rotation period, different limb darkening coefficients, and different
photometric uncertainty, meaning we must factorize $M$ different covariance
matrices. Fortunately, the spherical harmonic covariance, $\pmb{\Sigma}_\mathbf{y}$
(\S\ref{sec:gp}) need only be computed once; this can then be linearly
transformed into the flux basis for each light curve. As the evaluation of
$\pmb{\Sigma}_\mathbf{y}$ is the computationally-intensive step, the
likelihood evaluation typically scales sub-linearly with $M$. Furthermore,
as we discussed in \ref{sec:other-marg}, it is possible to marginalize over
the period and limb darkening coefficients, which would remove the scaling
with $M$ entirely if the photometric precision were the same for all
light curves. Even if it is not, the algorithm can still be greatly sped up,
although an implementation of this is deferred to the next paper.
\xxx{DFM: Is this true? I.e., is there an efficient way to solve Equation~(\ref{eq:log-like})
    when $\pmb{\Sigma}$ is the same for all stars but $\mathbf{C}$ is different? Is there
    some limit in which an approximation of some sort can be made?}

@@

Our algorithm is also numerically stable over nearly all of the prior
volume up to $l_\mathrm{max} = 15$. Figure~\ref{fig:stability}
shows the log of the condition number of the covariance matrix in the
spherical harmonic basis, $\pmb{\Sigma}_\mathbf{y}$, as a function of the
spherical harmonic degree of the expansion for 100 draws from a uniform
prior over $r \in [10^\circ, 45^\circ]$, $\mu_\phi \in [0^\circ, 85^\circ]$,
$\sigma_\phi \in [5^\circ, 40^\circ]$, and $n \in [1, 50]$. The condition
number is nearly constant up to $l_\mathrm{max} = 15$ in almost all cases;
above this value, the algorithm suddenly becomes unstable and the
covariance is ill-conditioned. The instability occurs within the computation
of the latitude and longitude moment integrals (Appendix~\ref{sec:integrals})
and is likely due to the large number of operations involving linear
combinations of hypergeometric and gamma functions. While it may be possible to
achieve stability at higher values of $l_\mathrm{max}$ via careful
reparametrization of some of those equations, we find that $l_\mathrm{max} = 15$
is high enough for most practical purposes. We plan to revisit this point
in future work.

Finally, instabilities can also occur if $\sigma_\phi$ is too small and/or
$n$ is too large. Values of $\sigma_\phi \lesssim 1^\circ$ lead to instabilities
in the computation of the hypergeometric functions (Appendix~\ref{sec:lat}),
while values of $n \gtrsim 50$ can sometimes cause the Cholesky factorization
of the covariance to fail (although this can be mitigated by adding a small
quantity to the diagonal to ensure positive-semidefinitess). In cases where
the algorithm goes (very) unstable, the log-likelihood evaluation returns
$-\infty$: in other words, they are silently rejected by an implicit prior.
Fortunately, these cases are likely unphysical: in practice, there should always be some
finite amount of variance in the latitudes of spots, and stars with more than
50 spots are likely too spotted for individual spots to be discernible in the
first place. Instead, we are likely sensitive to \emph{groups} of spots, which
our GP is flexible enough to model.
